---
title: "Reordering boxplots"
author: "Doug Nychka"
date: "1/27/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First create a data frame with several categories as factors. 

```{r}
Z<- c(1:20, 31:40, 101:110)
tag<- rep( c("D","A","C","B"), rep(10,4))
df<- data.frame( Z=Z,tag= factor(tag))
head( df)
```
Here a boxplot of data grouped by the factor. This is the default order based on sorting the names (A through D)

```{r}
boxplot( Z~ tag, data = df)
```
Now reorder in reverse order ( D throug A)

```{r}
boxplot( Z~ tag, data = df, at=4:1)
```

How about ordering by the size of the mean? Here we need to get the ranks of the groups based on the mean value.

```{r}
groupMeans<- tapply( Z, tag, mean, data=df)
index<- rank( groupMeans)
boxplot( Z~ tag, data = df, at=index)
```

Sometimes in ordering by a numeric value it is useful to just locate the boxplots by those values. Here we just plot the boxplots at there mean values.
This saves a step in ranking and creates a different horizontal axis and so the box width needs to be adjusted to the x axis limits.
I added the axis on top to avoid collision with the group labels and increased the range based on the data to accommodate the first box. 



```{r}
groupMeans<- tapply( Z, tag, mean, data=df)
boxplot( Z~ tag, data = df, 
         at=groupMeans,
         boxwex=8, xlim=range(df$Z) )
axis(3) 
```





