---
title: "Lecture04FakeHWSolutions"
author: "Doug Nychka"
date: "1/17/2021"
output: pdf_document
---

```{r include=FALSE}
# include = FALSE means no output 
# this makes for a cleaner solution.
knitr::opts_chunk$set(echo = TRUE)
library( fields) # also load the fields library
```
<!-- BEGIN Comment
A copy of the problem for reference but not printed out
## Problem 1
Create the RNATest data set and ...
* (A) What are the median and  mean for these RNA data? 
  Which is larger?
* (B) Plot a histogram of these data and locate the mean and median by vertical lines. (You can use the fields package function ```xline``` to to add the lines.)
* (C) Do any of the values  seem to  be outliers, i.e. unusual, worthy of scrutiny?
END Comment   -->

# Problem 1 

## Solution to (A)

```{r echo=FALSE}
# echo here means do not show these commands
# they were given in the problem
covid <-
  read.csv(file = "CDPHE_COVID19_Wastewater_Dashboard_Data.csv", header = T)
boulder_covid <- covid[covid$Utility == "Boulder", ]

RNATest<- na.omit( boulder_covid$SARS_CoV_2_copies_L )
```

Having created the subset of data ( ```RNATest```)  following the homework
instructions. 
Here are the mean and median. 

```{r}
mean( RNATest)
median(RNATest)
```

We see that the mean is larger. 

## Solution to (B)

<!--  Play around with the figure size so it 
      looks nice ... 
-->

```{r fig.height=6, fig.width = 6 }
hist( RNATest, nclass=10, 
        xlab="SARS/COVID copies", main="")
# don't make a title here 
fields:::xline(  mean(  RNATest), col="blue",    lwd=4)
fields:::xline(  median(RNATest), col="magenta", lwd=4)
title("Boulder utility WW measurements \
 median (magenta)  mean (blue)")
```

<!-- Insert a page break so next answer is on a new page.
This is a command in the markdown language 
-->

\pagebreak

## Solution to (C)

```{r}
max( RNATest)
```

Based on the shape of this histogram I see many values clustered below about
*125,000* and some larger values beyond that. However, the maximum value,
( $\approx 281,000$ )   
  seems to be much larger relative to  the others and should be examined more closely. 

<!--  
The $ symbols calls/trigger LaTex to evaluate what is inside
Of course one needs to know that the  approximate symbo in
Latex is specified by \approx .
In general  use  $ to enclose any LaTeX formula or math 
expression to see it in R Markdown. 
 -->









